---
title: "ELMER.data: Supporting data for the ELMER package"
author: "Tiago Chedraoui Silva, Simon Coetzee, Lijing Yao, Peggy Farnham, Hui Shen, Peter Laird, Houtan Noushmehr, Ben Berman"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
      highlight: tango
      toc: yes
      fig_caption: yes     
      toc_depth: 3
      toc_float:
        collapsed: yes
      number_sections: true
    editor_options: 
      chunk_output_type: inline
references:
- id: ref1
  title: HOCOMOCO a comprehensive collection of human transcription factor binding sites models
  author: 
  - family: Kulakovskiy, Ivan V and Medvedeva, Yulia A and Schaefer, Ulf and Kasianov, Artem S and Vorontsov, Ilya E and Bajic, Vladimir B and Makeev, Vsevolod
    given:
  journal: Nucleic acids research
  volume: 41
  number: D1
  pages: D195--D202
  issued:
    year: 2013    
- id: heinz2010simple
  title: Simple combinations of lineage-determining transcription factors prime cis-regulatory elements required for macrophage and B cell identities
  author: 
  - family: Heinz, Sven and Benner, Christopher and Spann, Nathanael and Bertolino, Eric and Lin, Yin C and Laslo, Peter and Cheng, Jason X and Murre, Cornelis and Singh, Harinder and Glass, Christopher K
    given:
  journal: Molecular cell
  volume: 38
  number: 4
  pages: 576--589
  issued:
    year: 2010    
- id: ELMER
  title: Inferring regulatory element landscapes and transcription factor networks from cancer methylomes
  author: 
  - family: Yao, L., Shen, H., Laird, P. W., Farnham, P. J., & Berman, B. P. 
    given:
  journal: Genome biology
  volume: 16
  number: 1
  pages: 105
  issued:
    year: 2015 
- id: wingender2014tfclass
  title: TFClass a classification of human transcription factors and their rodent orthologs
  author: 
  - family: Wingender, E., Schoeps, T., Haubrock, M., & Dönitz, J
    given:
  journal: Nucleic acids research
  pages: gku1064
  issued:
    year: 2014 
- id: zhou2016comprehensive
  title: Comprehensive characterization, annotation and innovative use of Infinium DNA methylation BeadChip probes
  author: 
  - family: Zhou, Wanding and Laird, Peter W and Shen, Hui
    given:
  journal: Nucleic Acids Research
  pages: gkw967
  issued:
    year: 2016
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{ELMER.data: Supporting data for the ELMER package}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---   

```{r, echo = FALSE,hide=TRUE, message=FALSE,warning=FALSE}
library(ELMER.data)
library(DT)
library(dplyr)
```

# Introduction

This document provides an introduction of the `r BiocStyle::Biocpkg("ELMER.data")`, which contains 
supporting data for `r BiocStyle::Biocpkg("ELMER")` [@ELMER]. `r BiocStyle::Biocpkg("ELMER")` is package using DNA methylation to 
identify enhancers, and correlates enhancer state with expression of nearby genes 
to identify one or more transcriptional targets. Transcription factor (TF) binding 
site analysis of enhancers is coupled with expression analysis of all TFs to 
infer upstream regulators. `r BiocStyle::Biocpkg("ELMER.data")` provide 3 necessary data for 
`r BiocStyle::Biocpkg("ELMER")` analysis:

1. Probes information: files with DNA methylation platforms metadata  retrieved from [http://zwdzwd.github.io/InfiniumAnnotation](http://zwdzwd.github.io/InfiniumAnnotation) [@zhou2016comprehensive].
2. Probes.motif: motif occurences within $\pm 250bp$ of probe sites on HM450K/EPIC array aligned against hg19/hg38.

## Installing and loading ELMER.data

To install this package, start R and enter

```{r, eval = FALSE}
devtools::install_github(repo = "tiagochst/ELMER.data")
library("ELMER.data")
library("GenomicRanges")
```

# Contents

## Probes information

Probes information were retrieved from [http://zwdzwd.github.io/InfiniumAnnotation](http://zwdzwd.github.io/InfiniumAnnotation) [@zhou2016comprehensive].

```{r, eval=FALSE, include=TRUE}
for(plat in c("450K","EPIC")) {
  for(genome in c("hg38","hg19")) {
    base <- "http://zwdzwd.io/InfiniumAnnotation/current/"
    if(plat == "EPIC") {
      annotation <- paste0(base,"EPIC/EPIC.manifest.rda")
    } else {
      annotation <- paste0(base,"hm450/hm450.manifest.rda")
    }
    if(genome == "hg38") annotation <- gsub(".rda",".hg38.rda", annotation)
    if(!file.exists(basename(annotation))) {
      if(Sys.info()["sysname"] == "Windows") mode <- "wb" else  mode <- "w"
      downloader::download(annotation, basename(annotation), mode = mode)
    }
  }
}
```
```{r, message=FALSE}
data("EPIC.manifest")
as.data.frame(EPIC.manifest)[1:5,] %>% datatable(options = list(scrollX = TRUE,pageLength = 5)) 
data("EPIC.manifest.hg38")
as.data.frame(EPIC.manifest.hg38)[1:5,] %>% datatable(options = list(scrollX = TRUE,pageLength = 5)) 
data("hm450.manifest.hg38")
as.data.frame(hm450.manifest.hg38)[1:5,] %>% datatable(options = list(scrollX = TRUE,pageLength = 5)) 
data("hm450.manifest")
as.data.frame(hm450.manifest)[1:5,] %>% datatable(options = list(scrollX = TRUE,pageLength = 5)) 
```

## Probes.motif

Probes.motif provides information for motif occurences within$\pm 250bp$ of probe 
sites on HM450K/EPIC array. HOMER [@heinz2010simple] was used with a p-value < 0.0001 to scan a $\pm 250bp$ 
region around each probe on HM450K/EPIC using HOCOMOCO V10 motif position weight 
matrices (PWMs) which provides transcription factor (TF) binding models for more than 600 human TFs.
This data set is used in get.enriched.motif function in `r BiocStyle::Biocpkg("ELMER")` to calculate 
Odds Ratio of motif enrichments for a given set of probes. This data is storaged 
in a sparse matrix with wuth 640 columns, there is one matrix for HM450K aligned to hg19, one for HM450K  aligned to hg38,
one for EPIC aligned to hg19, one for EPIC  aligned to hg38. Each row is each probe regions (annotation of the regions used can be found in 
[this repository](http://zwdzwd.github.io/InfiniumAnnotation)) and each 
column is motif from [http://hocomoco.autosome.ru/](HOCOMOCO) [@ref1]. The value 1 indicates the occurrence 
of a motif in a particular probe and 0 means no occurrence.

```{r}
data("Probes.motif.hg19.450K")
dim(Probes.motif.hg19.450K)
str(Probes.motif.hg19.450K)
```

```{r}
data("Probes.motif.hg38.450K")
dim(Probes.motif.hg38.450K)
str(Probes.motif.hg38.450K)
```

```{r}
data("Probes.motif.hg19.EPIC")
dim(Probes.motif.hg19.EPIC)
str(Probes.motif.hg19.EPIC)
```

```{r}
data("Probes.motif.hg38.EPIC")
dim(Probes.motif.hg38.EPIC)
str(Probes.motif.hg38.EPIC)
```

The following code was used to create the objects:
```{r, eval=FALSE, include=TRUE}
getInfiniumAnnotation <- function(plat = "450K", genome = "hg38"){
  message("Loading object: ",file)
  newenv <- new.env()
  if(plat == "EPIC" & genome == "hg19") data("EPIC.manifest", package = "ELMER.data",envir=newenv)
  if(plat == "EPIC" & genome == "hg38") data("EPIC.manifest.hg38", package = "ELMER.data",envir=newenv)
  if(plat == "450K" & genome == "hg19") data("hm450.manifest", package = "ELMER.data",envir=newenv)
  if(plat == "450K" & genome == "hg38") data("hm450.manifest.hg38", package = "ELMER.data",envir=newenv)
  annotation <- get(ls(newenv)[1],envir=newenv)   
  return(annotation)  
}
# To find for each probe the know motif we will use HOMER software (http://homer.salk.edu/homer/)
# Step:
# 1 - get DNA methylation probes annotation with the regions
# 2 - Make a bed file from it
# 3 - Execute section: Finding Instance of Specific Motifs from http://homer.salk.edu/homer/ngs/peakMotifs.html to the HOCOMOCO TF motifs
# Also, As HOMER is using more RAM than the available we will split the files in to 100k probes.
# Obs: for each probe we create a winddow of 500 bp (-size 500) around it. This might lead to false positives, but will not have false negatives.
# The false posives will be removed latter with some statistical tests.
TFBS.motif <- "http://hocomoco11.autosome.ru/final_bundle/hocomoco11/full/HUMAN/mono/HOCOMOCOv11_full_HUMAN_mono_homer_format_0.0001.motif"
if(!file.exists(basename(TFBS.motif))) downloader::download(TFBS.motif,basename(TFBS.motif))
for(plat in c("450K","EPIC")){
  for(gen in c("hg19","hg38")){
    
    file <- paste0(plat,gen,".txt")
    print(file)
    if(!file.exists(file)){
      # STEP 1
      gr <- getInfiniumAnnotation(plat = plat,genome =  gen)
      
      # This will remove masked probes. They have poor quality and might be arbitrarily positioned (Wanding Zhou)
      print(table(gr$MASK.general))
      gr <- gr[!gr$MASK.general]
      print(table(gr$MASK.general))
      
      df <- data.frame(seqnames=seqnames(gr),
                       starts=as.integer(start(gr)),
                       ends=end(gr),
                       names=names(gr),
                       scores=c(rep(".", length(gr))),
                       strands=strand(gr))
      step <- 50000 # nb of lines in each file 50K was selected to not explode RAM
      n <- nrow(df)
      for(j in 0:floor(n/step)){
        # STEP 2
        file.aux <- paste0(plat,gen,"_",j,".bed")
        if(!file.exists(gsub(".bed",".txt",file.aux))){
          end <- ifelse(((j + 1) * step) > n, n,((j + 1) * step))
          write.table(df[((j * step) + 1):end,], file = file.aux, col.names = F, quote = F,row.names = F,sep = "\t")
          
          # STEP 3
          cmd <- paste("source ~/.bash_rc; annotatePeaks.pl" ,file.aux, gen, "-m", basename(TFBS.motif), "-size 500 -cpu 12 >", gsub(".bed",".txt",file.aux))
          system(cmd)
        }
      }
      # We will merge the results from each file into one
      peaks <- NULL
      for(j in 0:floor(n/step)){
        aux <-  readr::read_tsv(paste0(plat,gen,"_",j,".txt"))
        colnames(aux)[1] <- "PeakID"
        if(is.null(peaks)) {
          peaks <- aux
        } else {
          peaks <- rbind(peaks, aux)
        }
      }
      readr::write_tsv(peaks,path=file,col_names = TRUE)
      print("DONE!")
      gc()
    }
  }
}

getMatrix <- function(filename) {
  motifs <- readr::read_tsv(file)
  # From 1 to 21 we have annotations then we have 640 motifs
  matrix <- Matrix::Matrix(0, nrow = nrow(motifs), ncol = ncol(motifs) - 21 ,sparse = TRUE)
  colnames(matrix) <- gsub(" Distance From Peak\\(sequence,strand,conservation\\)","",colnames(motifs)[-c(1:21)])
  rownames(matrix) <- motifs$PeakID
  matrix[!is.na(motifs[,-c(1:21)])] <- 1
  matrix <- as(matrix, "nsparseMatrix")
  return(matrix)
}

# This code will read the table with the motifs, save it as a sparce matrix
# and save all as a .rda that will be placed in ELMER.data
for(plat in c("450K","EPIC")){
  for(gen in c("hg19","hg38")){
    file <- paste0(plat,gen,".txt")
    if(file == "450Khg19.txt"){
      Probes.motif.hg19.450K <- getMatrix(file)
      save(Probes.motif.hg19.450K, file = "Probes.motif.hg19.450K.rda", compress = "xz")
      rm(Probes.motif.hg19.450K)
    } 
    if(file == "450Khg38.txt"){
      Probes.motif.hg38.450K <- getMatrix(file)
      save(Probes.motif.hg38.450K, file = "Probes.motif.hg38.450K.rda", compress = "xz")
      rm(Probes.motif.hg38.450K)
    }
    
    if(file == "EPIChg19.txt"){
      Probes.motif.hg19.EPIC <- getMatrix(file)
      save(Probes.motif.hg19.EPIC, file = "Probes.motif.hg19.EPIC.rda", compress = "xz")
      rm(Probes.motif.hg19.EPIC)
    }
    
    if(file == "EPIChg38.txt"){
      Probes.motif.hg38.EPIC <- getMatrix(file)
      save(Probes.motif.hg38.EPIC, file = "Probes.motif.hg38.EPIC.rda", compress = "xz")
      rm(Probes.motif.hg38.EPIC)
    }
  }
}
```

```{r}
data("Probes.motif.hg19.450K")
as.data.frame(as.matrix(Probes.motif.hg19.450K[1:20,1:20])) %>% 
  datatable(options = list(scrollX = TRUE,pageLength = 5)) 
```

<div class="panel panel-info">
<div class="panel-heading">Homer versions</div>
<div class="panel-body">
- Software: v4.9.1
- Genome hg19: v5.10
- Genome hg38: v5.10
</div>
</div>

# Session Information
******
```{r sessionInfo}
sessionInfo()
```

# References
